---
title: "Getting started with ICONS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with ICONS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 3.4, dpi = 96)
```

```{r setup}
library(ICONS)
```

## The problem

Confirmatory factor analysis is attractive: a sparse loading matrix, one factor
per group of variables, arbitrary covariance between factors. It has two
practical problems. You need to know the grouping in advance, and standard
implementations do not scale much past a few hundred variables.

High-throughput biomedical data usually violates the first requirement and
always violates the second. What it does offer is structure: covariance
matrices from genomics, proteomics, metabolomics and neuroimaging tend to show
*interconnected community structure* — groups of mutually correlated features,
the groups themselves correlated, and a residue of features belonging to no
group.

ICONS turns that structure into the missing prior knowledge. Communities are
detected from the covariance matrix, each becomes a factor, and the resulting
model has closed-form estimators throughout.

## A worked example

```{r}
data(sim)
dim(sim)
W <- cor(sim)
```

The structure is invisible in the variables' original order and obvious once
they are reordered — which is the whole point of the detection step.

```{r, fig.height = 3.4}
plot_matrix(W, main = "Original order")
```

### Detecting communities

`icons_detect()` repeatedly extracts the densest remaining subgraph. Each pass
strips the lowest-degree node and scores the survivors by the generalised
density $w(S)/|S|^{2\lambda}$; the best-scoring set becomes a community and the
process restarts on what is left.

```{r}
part <- icons_detect(W, threshold = 0.6)
part
```

Two settings govern this. `threshold` decides which edges are visible at all,
and `lambda` trades size against density — larger values give smaller, denser
communities. Variables never assigned anywhere become the **singleton set**,
reported separately rather than being treated as a community.

```{r}
plot_matrix(reorder_matrix(W, part), partition = part, main = "Reordered")
```

If reasoning about a threshold on the correlation scale is awkward, give a
quantile of the observed edge weights instead:

```{r}
icons_detect(W, probs = 0.95)$sizes
```

### Letting the data choose the settings

```{r}
tuned <- icons_tune(W, sim,
                    probs  = seq(0.90, 0.98, by = 0.02),
                    lambda = c(0.4, 0.6, 0.8))
tuned
```

```{r, fig.height = 4}
plot(tuned)
```

The surface matters more than the argmin. The criterion is evaluated on the
same matrix that selected the parameters, so the minimum is optimistically
biased; and if the lines are flat, the data are not really distinguishing the
candidates. Read the plot before quoting the selected values.

Note also how much more the threshold moves the criterion than `lambda` does.
That is typical.

### How many factors?

`n_factors()` admits communities one at a time, largest first, and records how
much covariance is left unexplained.

```{r, fig.height = 4}
nf <- n_factors(sim, part)
nf
plot(nf)
```

The whole path costs one pass over the data plus $O(K^3)$, not $K$ separate
fits: $\hat a_{kk}$ and $\hat b_{kk'}$ depend only on their own blocks, so
admitting more communities never changes the ones already in.

Here the curve flattens after the eighth community — the ninth is the
low-correlation residue visible in the bottom-right of the heat map.

### Fitting the model

```{r}
fit <- scfa(sim, part)
fit
```

With variables ordered by community the model is

$$X = Lf + u, \qquad \Sigma = L\Sigma_f L^\top + \Sigma_u,$$

with $L = \mathrm{Bdiag}(1_{p_1}, \dots, 1_{p_K})$, $\Sigma_f = (b_{kk'})$ and
$\Sigma_u = \mathrm{Bdiag}(a_{11}I_{p_1}, \dots, a_{KK}I_{p_K})$. Every
parameter has a closed form:

$$\hat a_{kk} = \frac{p_k \operatorname{tr}(S_{kk}) - \operatorname{sum}(S_{kk})}{p_k(p_k-1)},
\qquad
\hat b_{kk'} = \frac{\operatorname{sum}(S_{kk'})}{p_k p_{k'}} \quad (k \neq k').$$

These are uniformly minimum-variance unbiased whenever $K + K(K+1)/2 < n$.

```{r}
round(fit$Sigma_f[1:4, 1:4], 3)   # factor covariance
round(fit$a[1:4], 3)              # error variance per community
dim(fit$scores)                   # factor scores
```

The factor scores are just the community means of the centred data. That is not
a shortcut: Theorem 2 of Yang et al. (2024) shows the OLS, GLS and feasible GLS
estimators all coincide under this $\Sigma_u$.

### Inference

The paper also derives exact variances, so Wald intervals need no bootstrap.

```{r}
head(confint(fit), 6)
```

```{r}
summary(fit)
```

`coef()`, `vcov()`, `fitted()`, `residuals()`, `predict()` and `nobs()` behave
as you would expect, and `factor_loadings()` builds $\hat L$ on demand.

```{r}
dim(factor_loadings(fit))
head(sigma_u(fit))
```

## Why it scales

The estimators above depend on the sample covariance matrix $S$ only through
$\operatorname{sum}(S_{kk'})$ and $\operatorname{tr}(S_{kk})$. Both follow from
a single pass over the data: the within-community sums $T = X_c L$, which is
$n \times K$, and the per-variable sums of squares.

So $S$ is never formed. `scfa()` costs $O(np + nK^2)$ rather than $O(np^2)$,
and its memory is $O(nK + p)$ rather than $O(p^2)$. Fit criteria use the
$n \times n$ Gram matrix, giving $O(n^2 p)$, which is the right way round when
$p \gg n$.

In practice a fit at $p = 50{,}000$ takes a few seconds. Forming the covariance
matrix alone at that size would need 20 GB.

Detection is compiled: the thresholded matrix is held once in compressed sparse
row form and the adaptive loop runs entirely in C++, so no shrinking submatrix
is ever copied.

## Migrating from 0.1.x

Old names still work and warn once per session.

| 0.1.x | 0.2.0 |
| --- | --- |
| `dense()` | `icons_detect()` |
| `param_tuning_sigmau()` | `icons_tune()` |
| `k.elbow()` | `n_factors()` |
| `plotMatrix()` | `plot_matrix()` |
| `get_membership()` | `as_membership()` |
| `get_index()` | `block_index()` |
| `get_vectorform()` | `half_vec()` |

`scfa()` keeps its name but takes a partition object or a membership vector in
place of the `CID`/`Clist` pair:

```{r, eval = FALSE}
# 0.1.x
res <- dense(W, 0.6, 0.5)
fa  <- scfa(sim, res$CID, res$Clist)

# 0.2.0
part <- icons_detect(W, threshold = 0.6, lambda = 0.5)
fit  <- scfa(sim, part)
```

Results will differ, deliberately. 0.1.x reported `cov(F_hat)` as the factor
covariance; Theorem 3 gives
$\operatorname{cov}(\hat f) = \Sigma_f + \operatorname{diag}(a_{kk}/p_k)$
exactly, so that estimator is biased upward on the diagonal. You can see the
identity directly:

```{r}
round(diag(cov(fit$scores)) - (diag(fit$Sigma_f) + fit$a / fit$sizes), 12)
```

To reproduce numbers published with 0.1.x, use `criterion = "legacy"`, which
recreates the old tuning objective exactly.

## References

Yang, Y., Ma, T., Bi, C., & Chen, S. (2024). Semi-confirmatory factor analysis
for high-dimensional data with interconnected community structures.
*arXiv:2401.00624*.

Yang, Y., Chen, C., & Chen, S. (2024). Covariance matrix estimation for
high-throughput biomedical data with interconnected communities.
*The American Statistician*, 78(4), 401–411.

Chen, S., Zhang, Y., Wu, Q., Bi, C., Kochunov, P., & Hong, L. E. (2024).
Identifying covariate-related subnetworks for whole-brain connectome analysis.
*Biostatistics*, 25(2), 541–558.

Wu, Q., Huang, X., Culbreth, A. J., Waltz, J. A., Hong, L. E., & Chen, S.
(2022). Extracting brain disease-related connectome subgraphs by adaptive dense
subgraph discovery. *Biometrics*, 78(4), 1566–1578.
